# Conditional Random Intercept Models

**Data for this chapter:** `lq2002.csv`

This week, we are going to use data from Gavin and Hofmann (2002), a study on organizational climate and attitudes published in *Leadership Quarterly*. Here, we have individual soldiers nested within companies. This is the same dataset that Garson uses in Chapter 6, so you can recreate his analysis.

The structure is the same as students within schools, which is a useful reminder that multilevel modeling is not about education specifically. Anywhere individuals are grouped, these methods apply.

## Load some packages to help with the analysis and data management

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(lme4)
```

## Load and explore the data

```{r}
#| message: false

lq2002 <- read_csv("lq2002.csv")

glimpse(lq2002)
```

Remember our old friend `describe` from the `psych` package a few weeks ago? This is a great way to get a detailed summary of all the variables in a dataset.

```{r}
psych::describe(lq2002,
                fast = TRUE)
```

Writing `psych::describe()` rather than just `describe()` is a small habit worth adopting. Several packages define a function by that name, and the `package::function()` form says exactly which one you mean. If you have `Hmisc` loaded, as we did last week, the unqualified version may not be the one you expect.

The variables we care about are `HOSTILE` (a hostility scale score, our outcome), `TSIG` (task significance), `LEAD` (leadership climate), and `COMPID` (company, our clustering variable). The variables beginning with `G` are group-level aggregates, which matter later in this chapter.

## Verifying a group-level aggregate by hand

`GTSIG`, the company average of task significance, is already sitting in the data as a column. Before you trust it — or build your own version of it on a dataset that doesn't hand you one — it's worth seeing exactly what it is: a mean, computed within each company and copied back onto every row in that company.

```{r}
lq2002 <- lq2002 %>%
  group_by(COMPID) %>%
  mutate(gtsig_check = mean(TSIG, na.rm = TRUE)) %>%
  ungroup()

lq2002 %>%
  select(COMPID, TSIG, GTSIG, gtsig_check) %>%
  distinct(COMPID, GTSIG, gtsig_check) %>%
  head()
```

`group_by()` tells R which rows belong together, `mutate()` computes the mean within each group and writes it back onto every row in that group, and `ungroup()` removes the grouping so later steps don't keep operating group-by-group without your noticing. `gtsig_check` matches `GTSIG` to within rounding error — confirmation that a group-level aggregate is nothing more exotic than this.

## Visualizing the key variables

Let's take a look at some of the key variables for our analysis. We have three variables, all of which are scale scores constructed just like the example above.

```{r}
#| label: fig-key-vars
#| fig-cap: "Distributions of the three key variables."
#| message: false

lq2002 %>%
  select(HOSTILE, TSIG, LEAD) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  mutate(variable = factor(variable,
                            levels = c("HOSTILE", "TSIG", "LEAD"),
                            labels = c("Hostility", "Task Significance", "Leadership Climate"))) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 20, fill = "grey70", color = "white") +
  facet_wrap(~ variable, scales = "free_x", nrow = 1) +
  labs(x = NULL, y = "Count",
       caption = "N = 2,042 U.S. Army active duty soldiers. Data from Gavin and Hofmann (2002).") +
  theme_minimal()
```

Notice the shape of the hostility distribution in the leftmost panel. It piles up near the low end, which is what you would expect for a measure where most people report very little of the thing. That floor will matter when we interpret coefficients later, because there is limited room to move downward.

## Running the multilevel models

### Step one: the null model

```{r}
model.0 <- lmer(HOSTILE ~ (1|COMPID), REML = FALSE, data = lq2002)

summary(model.0)
```

The `(1|COMPID)` term is what makes this multilevel. Everything else follows from it.

#### Post-estimation task: calculate ICC

Here's a quick and easy way to get our friend the ICC!

```{r}
null.ICC <- 0.05765/(0.05765 + 1.01668)
null.ICC
```

Note how much smaller this is than the ICC we found for schools and math achievement in the last chapter. That is substantively interesting rather than a problem. Soldiers' hostility is mostly an individual matter, with companies accounting for a modest share of the variation.

A small ICC does not mean you should abandon multilevel modeling. It still corrects the standard errors, and it still lets you ask questions about company-level predictors, which is exactly where this chapter is heading.

Doing that arithmetic by hand once is worth it, because it shows you exactly what an ICC is: a ratio of variance components. Going forward, `performance::icc()` will do it for you directly from the fitted model, which is faster and less error-prone once you trust the concept.

```{r}
performance::icc(model.0)
```

#### Using `lmerTest` to evaluate random effects

```{r}
lmerTest::ranova(model.0)
```

This performs a likelihood ratio test comparing the model with and without the company random effect — the same comparison the ICC hinted at above, now made formal. A significant result means companies differ enough on hostility that the clustering is worth modeling explicitly.

### Step two: adding a level-one predictor, `TSIG`

```{r}
model.1 <- lmer(HOSTILE ~ TSIG + (1|COMPID), REML = FALSE, data = lq2002)

summary(model.1)
```

The coefficient for `TSIG` describes how hostility differs between two soldiers who differ by one unit in task significance. Because `TSIG` varies both within companies and between them, this single estimate blends two different comparisons. We separate them in step five.

### Step three: adding an additional level-one predictor, `LEAD`

```{r}
model.2 <- lmer(HOSTILE ~ TSIG + LEAD + (1|COMPID), REML = FALSE, data = lq2002)

summary(model.2)
```

Watch what happens to the `TSIG` coefficient when `LEAD` enters. Soldiers who find their work meaningful also tend to report better leadership climate, so some of what model 1 attributed to task significance was partly leadership.

### Centering before testing an interaction

`TSIG` and `LEAD` are both scored so that 0 is outside the range anyone actually answered — nobody scores a 0 on a 1–5 scale. Left uncentered, the main effects in an interaction model describe soldiers at that nonexistent zero point, which is not a useful thing to describe. Centering first fixes that.

```{r}
lq2002 <- lq2002 %>%
  mutate(TSIG_c = TSIG - mean(TSIG, na.rm = TRUE),
         LEAD_c = LEAD - mean(LEAD, na.rm = TRUE))
```

This is **grand-mean centering**: every soldier's score is expressed as a distance from the one overall sample mean, so the intercept in a model built from these centered variables now describes a soldier at the *average* level of both predictors, not at zero. Companies keep whatever between-company differences they had before centering — everyone in a company shifts by the same constant, so the centered variable still carries the full mix of within- and between-company variation that `TSIG` always had.

That is different from **group-mean centering**, which you have already done without naming it: `TSIG - gtsig_check`, subtracting each soldier's own company average rather than the grand average, would zero out every company's mean and leave only within-company variation behind. Group-mean centering answers "how does this soldier compare to their own company?" Grand-mean centering answers "how does this soldier compare to everyone?" We want the grand-mean version here, for a within-level interaction. Step five asks the group-mean question instead — without literally centering, it turns out that keeping `TSIG` on its raw scale and adding `GTSIG` alongside it isolates the same within-company effect that group-mean centering would.

### Step four: testing the interaction of `LEAD` and `TSIG`

```{r}
model.3 <- lmer(HOSTILE ~ TSIG_c + LEAD_c + TSIG_c:LEAD_c + (1|COMPID), REML = FALSE, data = lq2002)

summary(model.3)
```

The `:` operator specifies an interaction. Writing `TSIG_c + LEAD_c + TSIG_c:LEAD_c` includes both main effects plus their interaction. The shorthand `TSIG_c*LEAD_c` does the same thing, and the longer form makes the structure visible, which is why it is worth writing out while you are learning.

Centering changed the main effects and the intercept — they now describe an average soldier instead of a nonexistent one — but the interaction coefficient itself did not move. That is not a coincidence: centering only ever changes what the main effects mean, never whether or how strongly two predictors interact.

An interaction here asks whether the relationship between task significance and hostility depends on leadership climate. Both predictors are at level 1, so this is a *within-level* interaction. In the next chapter we look at cross-level interactions, where a level-2 variable moderates a level-1 slope.

### Step five: adding a level-two predictor, `GTSIG`

```{r}
model.4 <- lmer(HOSTILE ~ TSIG + LEAD + GTSIG + (1|COMPID), REML = FALSE, data = lq2002)

summary(model.4)
```

`GTSIG` is the company average of task significance. Having both `TSIG` and `GTSIG` in the model is not redundancy, even though it looks that way at first. The two coefficients answer different questions.

The individual-level coefficient asks: within a given company, do soldiers who report higher task significance report less hostility? The group-level coefficient asks: do companies where task significance is high on average have lower hostility overall? A soldier can be high on task significance in a company where everyone is low, or low in a company where everyone is high. Those are genuinely different situations and can have different consequences.

### Step six: adding a second level-two predictor, `GLEAD`

```{r}
model.5 <- lmer(HOSTILE ~ TSIG + LEAD + GTSIG + GLEAD + (1|COMPID), REML = FALSE, data = lq2002)

summary(model.5)
```

Track the company-level variance as you move from the null model through model 5. It should shrink as level-2 predictors enter, because those predictors are explaining between-company differences the null model could only describe. That shrinkage is the clearest signal that a level-2 predictor is doing real work.

## Should we include that interaction? Comparing `model.2` with `model.3`

```{r}
anova(model.2, model.3)
```

`anova()` on two `lmer` objects performs a likelihood ratio test. It is valid here because both models were fit with `REML = FALSE`, that is, with maximum likelihood. Comparing REML-fitted models that differ in their fixed effects is not valid, which is one reason we use ML throughout this book.

The LRT above lands right at the edge of conventional significance. That is exactly the situation where looking at the interaction, rather than just its p-value, earns its keep.

## Visualizing the interaction

```{r}
#| label: fig-interaction
#| fig-cap: "Predicted hostility by task significance, at three levels of leadership climate (centered scales)."
#| message: false

library(ggeffects)

plot(ggpredict(model.3, terms = c("TSIG_c", "LEAD_c [meansd]")))
```

`terms = c("TSIG_c", "LEAD_c [meansd]")` puts centered `TSIG_c` on the x-axis and draws one line for each of three representative values of `LEAD_c`: one standard deviation below its mean, at its mean, and one standard deviation above — 0 on this centered scale is an average soldier, not a nonexistent one. The three lines are not quite parallel — the relationship between task significance and hostility is a little steeper under poor leadership climate than under good leadership climate — which is the same thing the borderline LRT is telling you, just easier to see.

## Using `modelsummary` and `broom.mixed` to organize your results

```{r}
#| message: false

library(modelsummary)
library(broom.mixed)

models <- list(model.0, model.1, model.2, model.3, model.4, model.5)

modelsummary(models)
```

Reporting six models one `summary()` at a time is unreadable. A single comparison table is how this belongs in a paper, and reading across the columns is how you see what each addition changed.

`broom.mixed` is what makes this work. `modelsummary` knows how to format tables, but it relies on `broom.mixed` to translate `lmer` objects into tidy data frames first. Load it even though you never call it directly.

## HTML version that you can open in Word

```{r}
#| eval: false

modelsummary(models, output = 'msum.html', title = 'MLM Estimates')
```

This writes the table to a file in your working directory. Word opens HTML tables directly, so you can generate this, open it in Word, and paste it into a manuscript. `modelsummary` will also write `.docx` directly if you prefer: change the extension to `msum.docx`.

## A note on your content review

Your Module 4 content review uses the Best in Class classroom data rather than these Army data. The steps are the same: null model, ICC, verify a group-level mean by hand the way you just did with `GTSIG`, add predictors, and decide whether a level-two aggregate earns its place beyond the individual-level version. What changes is the substance of the question, not the code.

## What to take from this chapter

Three ideas carry forward. Level-1 and level-2 versions of the same construct answer different questions and belong in the model together. Variance components tell a story as you build up models, and watching them shrink shows where your predictors are doing work. And a series of models belongs in one table, not six.

The next chapter relaxes an assumption we have been making all along: that the relationship between a predictor and the outcome is the same in every company.
