# Longitudinal and Growth Curve Models

**Data for this chapter:** `STAR_long.dta` and `STAR_wide.dta`

::: {.callout-note}
## Data files needed
Copy `STAR_long.dta` and `STAR_wide.dta` into this project folder before rendering.
:::

Here the nesting changes character. Instead of students within schools, we have *observations within students*: repeated measures over time. The mathematics barely changes. The interpretation changes completely, and it is worth slowing down for that shift.

These data follow more than 1,300 Project STAR students from 3rd through 8th grade, six observations each.

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(haven)
library(lme4)
library(lmerTest)
library(performance)
library(modelsummary)
library(ggeffects)
```

```{r}
star_long <- read_dta("STAR_long.dta")

glimpse(star_long)
```

## Coding time

Almost always, code time so that zero is a meaningful moment. Here grade runs 3 to 8, so we subtract 3 and get a variable running 0 to 5.

```{r}
star_long <- star_long |>
  mutate(grade_0 = grade - 3)
```

This is not cosmetic. The intercept in a growth model is the predicted outcome when every predictor equals zero, including time. Left as 3 through 8, the intercept would describe a hypothetical student in grade zero, extrapolated three years before anyone was measured. Centered at the first wave, the intercept becomes the predicted score at the study's start, which is a quantity you can defend in print.

## Looking at trajectories first

```{r}
#| label: fig-spaghetti
#| fig-cap: "Individual trajectories for 30 randomly selected students, with the overall average in teal."
#| message: false

set.seed(664)
all_ids <- unique(star_long$stdntid)
sampled_ids <- sample(all_ids, min(30, length(all_ids)))

star_long |>
  filter(stdntid %in% sampled_ids) |>
  ggplot(aes(x = grade_0, y = science, group = stdntid)) +
  geom_line(alpha = 0.35) +
  geom_smooth(aes(group = NULL), method = "loess", color = "#1a7a6e", linewidth = 1.2) +
  labs(title = "Science Score Trajectories",
       subtitle = "30 randomly selected students",
       x = "Grades since 3rd", y = "Science score") +
  theme_minimal()
```

Look at this before fitting anything. Do students start in different places? Do they rise at different rates? Does the average curve bend? Those three questions map onto the random intercept, the random slope, and the polynomial terms we are about to test.

## Null model versus null growth model

```{r}
m_null <- lmer(science ~ 1 + (1 | stdntid), data = star_long, REML = FALSE)

summary(m_null)
```

```{r}
m_growth_null <- lmer(science ~ grade_0 + (1 | stdntid), data = star_long, REML = FALSE)

summary(m_growth_null)
```

Compare the residual variance across the two models. It drops substantially once time is in the model, and the reason is worth stating carefully. In the first model, "within-student variance" lumps together everything that makes a student's six scores differ, including the fact that they are growing. In the second, growth has been pulled out into a fixed effect, so the residual now represents deviation from a student's own trajectory rather than deviation from their own average.

This is the conceptual heart of growth modeling. The level-one units are occasions, and what varies at level one is how far a particular measurement sits from where the model expected that student to be at that time.

## A conditional growth model

```{r}
star_long <- star_long |>
  mutate(
    gender_f = as_factor(gender),
    race_f   = as_factor(race)
  )

m_cond <- lmer(science ~ grade_0 + gender_f + race_f + ac_mot + (1 | stdntid),
               data = star_long, REML = FALSE)

summary(m_cond)
```

Gender, race, and academic motivation are measured once per student, so they are level-two predictors even though they appear on every row of the long dataset. They can shift a student's whole trajectory up or down, but on their own they cannot change its slope. For that we need an interaction with time.

## Does motivation predict growth?

```{r}
m_interact <- lmer(science ~ grade_0 * ac_mot + gender_f + race_f + (1 | stdntid),
                   data = star_long, REML = FALSE)

summary(m_interact)
```

The research question, in plain language: do students who start out more academically motivated gain science achievement at a different rate than students who start out less motivated? The main effect of `ac_mot` is a difference in starting point. The interaction is a difference in slope. Those are separate claims, and conflating them is one of the most common errors in write-ups of growth models.

```{r}
#| label: fig-motivation
#| fig-cap: "Predicted science trajectories at low, average, and high academic motivation."

plot(ggpredict(m_interact, terms = c("grade_0", "ac_mot")))
```

## A random slope for time

```{r}
m_rs <- lmer(science ~ grade_0 * ac_mot + gender_f + race_f +
               (1 + grade_0 | stdntid),
             data = star_long, REML = FALSE)

summary(m_rs)
```

```{r}
ranova(m_rs)
```

This random slope says students differ not just in where they start but in how fast they climb. Picture two students with the same 3rd grade score: one gains steadily each year, the other plateaus after 5th grade. A random intercept model cannot tell them apart. This one can.

## Nonlinear growth

Growth is often not a straight line. Using the wide-format data, we reshape and test polynomial terms.

```{r}
star_wide <- read_dta("STAR_wide.dta")

star_clean <- star_wide |>
  pivot_longer(
    cols = starts_with("science"),
    names_to = "grade",
    values_to = "science",
    names_prefix = "science"
  ) |>
  mutate(
    grade = as.numeric(grade),
    grade0 = grade - 3,
    grade0_sq = grade0^2,
    grade0_cub = grade0^3
  )

glimpse(star_clean)
```

A sanity check after any reshape. Each student should now appear six times:

```{r}
star_clean |>
  count(stdntid) |>
  count(n, name = "n_students")
```

```{r}
m_lin  <- lmer(science ~ grade0 + (1 | stdntid), data = star_clean, REML = FALSE)
m_quad <- lmer(science ~ grade0 + grade0_sq + (1 | stdntid), data = star_clean, REML = FALSE)
m_cub  <- lmer(science ~ grade0 + grade0_sq + grade0_cub + (1 | stdntid), data = star_clean, REML = FALSE)

anova(m_lin, m_quad, m_cub)
```

Note `REML = FALSE` here. When comparing models that differ in their fixed effects, maximum likelihood is the appropriate estimator; REML likelihoods are not comparable across different fixed-effect specifications. This is a detail that quietly invalidates a lot of published model comparisons.

```{r}
AIC(m_lin, m_quad, m_cub)
BIC(m_lin, m_quad, m_cub)
```

Interpret the sign of the highest-order term you keep. A negative quadratic means growth decelerates, with gains getting smaller in later grades. It does not mean scores decline, which is a misreading that appears constantly. Scores decline only if the curve actually turns over within the range of your data, and you should check the plot before claiming it.

## Heterogeneous variance structures

Standard multilevel models assume the level-one residual variance is constant across occasions. That assumption is often wrong for longitudinal data, where scores tend to spread out as children develop. The `nlme` package can relax it.

```{r}
#| message: false

library(nlme)

m_homog <- lme(science ~ grade0 + grade0_sq,
               random = ~ 1 | stdntid,
               data = star_clean, method = "ML",
               na.action = na.omit)

m_heterog <- lme(science ~ grade0 + grade0_sq,
                 random = ~ 1 | stdntid,
                 weights = varIdent(form = ~ 1 | grade0),
                 data = star_clean, method = "ML",
                 na.action = na.omit)

anova(m_homog, m_heterog)
```

```{r}
summary(m_heterog)$modelStruct$varStruct
```

In plain terms: the standard model assumes measurements are equally noisy at every grade. The heterogeneous structure lets 8th grade scores be more variable than 3rd grade scores, which matters whenever a construct is still developing or a test has floor or ceiling effects at some ages.

::: {.callout-note}
## Two packages, two conventions
`lme4::lmer()` and `nlme::lme()` fit overlapping but different families of models, with different syntax and slightly different likelihood computations. Small numerical differences between them are expected and do not indicate that either is wrong. Load `nlme` after `lme4` with care, since both define functions with similar names.
:::
