# Intro to Growth Models

**Data for this chapter:** `egmerged.dta`

::: {.callout-note}
## Data file needed
Copy `egmerged.dta` into this project folder before rendering.
:::

This week, we go to a dataset collected as part of the US Sustaining Effects Study (Carter 1984), which attempted to assess the sustaining impacts of compensatory schooling for students over time.

The nesting changes character here. Instead of students within schools, we have *observations within students*: repeated measures over time. The mathematics barely changes. The interpretation changes completely, and it is worth slowing down for that shift.

## Load in our MVP packages

```{r}
#| label: setup

suppressPackageStartupMessages(library(tidyverse))
library(lme4)
```

## Load in the data

```{r}
egmerged <- haven::read_dta("egmerged.dta")

glimpse(egmerged)
```

Notice the shape. There is one row per student *per year*, not one row per student. This is called long format, and it is what growth models require. A student who was measured six times appears in six rows, with their time-invariant characteristics repeated on each one.

## Explore the data and ID variables using `describe` from the `Hmisc` package

```{r}
Hmisc::describe(egmerged)
```

## A little recoding: create factor versions of categorical variables

Our model is easier to interpret if time starts at 0.

```{r}
egmerged.clean <- egmerged %>%
  mutate(.,
         retained.fac = as_factor(retained),
         female.fac = as_factor(female),
         black.fac = as_factor(black),
         hispanic.fac = as_factor(hispanic),
         year0 = year - 1)
```

That last line is doing more work than it appears to. The intercept in any regression is the predicted outcome when every predictor equals zero, including time. Left as `year` running from 1, the intercept would describe a hypothetical measurement occasion before the study began. Recoded so the first wave is 0, the intercept becomes the predicted score at the start of the study, which is a quantity you can defend in print.

## Use `arrange` to examine variables that vary by student, year, and school

```{r}
egmerged %>%
  arrange(., childid, year)
```

Sorting by student and then by year lets you read one child's trajectory across consecutive rows. Do this before modeling. You will see immediately whether every student has the same number of observations, whether years are consecutive, and whether the outcome moves in a plausible way.

## Summarize math score by year, to look for preliminary changes over time

```{r}
egmerged %>%
  group_by(year) %>%
  summarize(avg_math_score = mean(math, na.rm = TRUE)
            )
```

## Same thing, by grade: the only difference should be missing or retained students

```{r}
egmerged %>%
  group_by(grade) %>%
  summarize(avg_math_score = mean(math, na.rm = TRUE)
            )
```

Comparing these two tables is a nice diagnostic. Year counts time since the study started; grade counts progress through school. For most students they move together, so the two tables should look nearly identical. Where they diverge, you are seeing students who were retained and therefore fell out of step, which is exactly what the `retained` variable records.

## Create summary data frame of student-level variables

```{r}
student.level <-egmerged.clean %>%
  group_by(childid) %>%
  summarize(.,
            black = mean(black),
            hispanic = mean(hispanic),
            female = mean(female),
            retained = max(retained),
            math = mean(math, na.rm = TRUE))

glimpse(student.level)
```

This collapses the long dataset down to one row per student, which is what you need for a descriptive table. Note the different summary functions: `mean()` for characteristics that are constant within a student, so the mean just returns that constant; `max()` for `retained`, which flags a student who was ever retained; and `mean()` for math, which here gives each student's average across waves.

## Table 1 for student-level variables

```{r}
#| message: false

library(table1)

table1(~ math + retained + black + hispanic, data = student.level)
```

The `table1` package produces the descriptive table that opens nearly every quantitative paper. Building it from the collapsed student-level data rather than the long data matters: run it on the long file and every student would be counted once per year, inflating your N and misstating the proportions.

## Create summary data frame of school-level variables

```{r}
school.level <-egmerged.clean %>%
  group_by(schoolid) %>%
  summarize(.,
            lowinc = mean(lowinc),
            mobility = mean(mobility),
            size = mean(size))

glimpse(school.level)
```

## Table 1 for school-level variables

```{r}
library(table1)

table1(~ lowinc + mobility + size, data = school.level)
```

Same logic one level up. School characteristics get described across schools, not across student-years.

## Distribution of the outcome

```{r}
#| label: fig-density
#| fig-cap: "Distribution of math scores across all observations."

density(egmerged.clean$math) %>%
  plot(., main = "Distribution of Math Scores")
```

A density plot is a smoothed alternative to a histogram, useful when you care about the shape rather than the exact bin counts.

## Getting fancy: maybe plot over time?

`scatter.smooth` is a nice way to get a sense of the trend in the data. But keep in mind, this type of plot does NOT account for observations nested within students, so it should be treated as exploratory.

```{r}
#| label: fig-scatter
#| fig-cap: "Math scores over time, ignoring the nesting."

scatter.smooth(x = egmerged.clean$year,
               y = egmerged.clean$math,
               main = "Distribution of Math Scores, Over Time")
```

That caveat is the whole reason for the rest of this chapter. The smoother here treats every point as independent, so it describes the average trend across all observations without knowing that groups of them belong to the same child. It is a good first look and a bad final answer.

## Run a regular null model for math scores

```{r}
model.null <- lmer(math ~ (1|childid), data = egmerged.clean)

summary(model.null)
```

### Calculate ICC

```{r}
null.icc <- 0.8483/(0.8483 + 1.5247)
null.icc
```

Interpret this the way you would any ICC, but notice what the levels now mean. The between-cluster variance is between *students*, and the within-cluster variance is within a student across occasions. So the ICC answers: how much of the variation in math scores is stable differences between children, versus fluctuation within a child over time?

## Now, run a GROWTH null model for math scores (with year included)

```{r}
model.null.growth <- lmer(math ~ year0 + (1|childid), data = egmerged.clean)

summary(model.null.growth)
```

### Calculate GROWTH ICC

```{r}
null.growth.icc <- 0.8683/(0.8683 + 0.3470)
null.growth.icc
```

Compare the two residual variances: roughly 1.52 in the null model, roughly 0.35 once time is included. That is a large drop, and the reason is worth stating carefully.

In the first model, "within-student variance" lumped together everything that made a child's scores differ across occasions, including the fact that they were *growing*. In the second, growth has been pulled out into a fixed effect, so the residual now represents deviation from a student's own trajectory rather than deviation from their own average.

This is the conceptual heart of growth modeling. The level-1 units are occasions, and what varies at level 1 is how far a particular measurement sits from where the model expected that student to be at that time.

## Add a student-level predictor, female

```{r}
model.1 <- lmer(math ~ year0 + female + (1|childid), data = egmerged.clean)

summary(model.1)
```

`female` is measured once per student, so it is a level-2 predictor even though it appears on every row of the long dataset. It can shift a student's whole trajectory up or down, but on its own it cannot change the slope.

## Add predictors for race/ethnicity: this is just for the intercept

```{r}
model.2 <- lmer(math ~ year0 + female + black + hispanic + (1|childid), data = egmerged.clean)

summary(model.2)
```

The section heading is the important part. These coefficients describe differences in where students *start*, not differences in how fast they grow. Everything so far assumes a single common growth rate.

## Do students differ in their growth by race/ethnicity? This is an interaction effect

```{r}
model.3 <- lmer(math ~ year0 + female + black + hispanic + year0:black + year0:hispanic + (1|childid), data = egmerged.clean)

summary(model.3)
```

Interacting a student-level characteristic with time is how you ask whether trajectories diverge. The main effect is a difference in starting point; the interaction is a difference in slope. Those are separate claims, and conflating them is one of the most common errors in write-ups of growth models.

A gap that is constant over time and a gap that widens over time call for very different responses, and only the interaction can tell them apart.

## Add a random slope for year at the child level

Each child has their own slope.

```{r}
model.4 <- lmer(math ~ year0 + female + black + hispanic +  (year0|childid), data = egmerged.clean)

summary(model.4)
```

Note the difference from the previous model. The interaction asked whether *groups* of students grow at different rates. This asks whether *individual* students do, without trying to explain why. Picture two children with identical third-grade scores: one climbs steadily, the other plateaus. A random intercept model cannot tell them apart. This one can.

## Should we keep that random slope for time? Use `rand` to test

```{r}
lmerTest::rand(model.4)
```

## Put it all together: random slope, interactions (not for Latinx students)

```{r}
model.5 <- lmer(math ~ year0 + female + black + hispanic + year0:black + (year0|childid), data = egmerged.clean)

summary(model.5)
```

## Visualize our model by predicting scores

```{r}
predicted.vals <- broom.mixed::augment(model.5, effects = c("ran_pars", "fixed"), conf.int = TRUE)

glimpse(predicted.vals)
```

`augment()` returns the original data plus fitted values, residuals, and related quantities in a tidy data frame. Asking for both `"ran_pars"` and `"fixed"` means the fitted values incorporate each child's own random effects, not just the average trajectory, which is what makes the next plot show individual variation.

## And then plotting them

```{r}
#| label: fig-predicted
#| fig-cap: "Predicted math trajectories, Black students compared with all others."

ggplot(data = predicted.vals, mapping = aes(x = year0, y = .fitted, color = as_factor(black))) +
  geom_jitter(alpha = .50) +
  geom_smooth(method = "lm", formula = 'y ~ x') +
  labs(title = "Predicted Change in Math Scores Over Time",
      subtitle = "Black Students (Teal Line) vs. All Others (Orange Line)",
      caption = "Notes. N = 1,731 students in the Sustaining Effects Study (Carter, 1984). Year 0 = Kindergarten.") +
  ylab("Predicted Math Score") +
  xlab("Year") +
  theme(legend.position = "none")
```

The two lines are the interaction made visible. If they run parallel, the growth rates are the same and only the starting points differ. If they diverge, the gap is widening. Reading the coefficient tells you which; seeing the plot tells you how much it matters.

## Random slopes for a sample of 50 students

This gives a sense of the variability in slopes.

```{r}
#| label: fig-fifty
#| fig-cap: "Predicted trajectories for a subset of students."

predicted.vals50 <- predicted.vals %>%
  arrange(childid, year0) %>%
  slice(., 1:252)

ggplot(data = predicted.vals50, mapping = aes(x = year0, y = .fitted, color = as_factor(childid))) +
  geom_jitter(alpha = .50) +
  geom_smooth(method = "lm", formula = 'y ~ x') +
  labs(title = "Predicted Change in Math Scores Over Time",
      caption = "Notes. N = 1,731 students in the Sustaining Effects Study (Carter, 1984). Year 0 = Kindergarten.") +
  ylab("Predicted Math Score") +
  xlab("Year") +
  theme(legend.position = "none")
```

This is the picture the random slope variance is summarizing. Each line is one child. The spread of their angles *is* the slope variance, and seeing it directly is more convincing than any number in the output.

Slicing the first 252 rows is a quick way to grab roughly the first 50 students once the data are sorted, since most contribute several observations each.

## Using `modelsummary` and `broom.mixed` to organize your results

```{r}
#| message: false

library(modelsummary)
library(broom.mixed)

models <- list(model.null, model.null.growth, model.1, model.2, model.3, model.4, model.5)

modelsummary(models, output = "default")
```

## HTML version that you can open in Word

```{r}
#| eval: false

modelsummary(models, output = 'msum.html', title = 'MLM Estimates')
```

Next week we take this further, into growth that is not a straight line and into models that let the variance itself change over time.
