# Binary and Generalized Linear Mixed Models

**Data for this chapter:** `nsch_2018_topical.dta`

::: {.callout-note}
## Data file needed
Copy `nsch_2018_topical.dta` into this project folder before rendering.
:::

This week, we will look at the 2018 National Survey of Children's Health, available at <https://www.census.gov/data/datasets/2018/demo/nsch/nsch2018.html>.

Not every outcome is continuous. When the thing you are predicting is yes or no, the linear model is the wrong tool, and we move to generalized linear mixed models. Here the outcome is whether a child has asthma, and children are nested within states.

## Load in our MVP packages

```{r}
#| label: setup
#| message: false

suppressPackageStartupMessages(library(tidyverse))
library(lme4)
library(Hmisc)
```

## Part 1: loading in the data, and then recoding variables

### Load in the data

```{r}
nsch <- haven::read_dta("nsch_2018_topical.dta")

glimpse(nsch)
```

This is a large file with a great many variables, which is typical of national survey data. You will almost never use more than a handful, which is what the next step is for.

### The `select` function is really useful, especially with really big datasets

First, we narrow down the data by selecting only the variables we need. Then, we use the `as_factor` function to convert all variables to factors ahead of time.

```{r}
nsch.2 <- nsch %>%
  select(.,
         fipsst,
         hhid,
         k2q40a,
         k9q41,
         mold,
         physactiv) %>%
  as_factor()

glimpse(nsch.2)
```

Two things worth noting. Survey variables often carry uninformative names like `k2q40a`, which is why the codebook is essential and why we rename them shortly. And applying `as_factor()` to the whole data frame at once converts every labelled variable using the labels that came with the file, which saves doing it one at a time.

`fipsst` is the state FIPS code, and it is our clustering variable. `hhid` identifies households.

### The `table` function is a great way to figure out how variables are labelled

```{r}
table(nsch.2$k2q40a)
table(nsch.2$k9q41)
table(nsch.2$physactiv)
table(nsch.2$mold)
```

Do this before anything else. Large surveys encode several kinds of non-response as ordinary categories, and if you skip this step those categories quietly become levels in your model.

### Now, we can use `filter` to only keep observations that have valid categories

```{r}
nsch.2.clean <- nsch.2 %>%
  filter(!(k2q40a %in%
                  c("No valid response", "Suppressed for confidentiality", "Not in universe", "Logical skip"))) %>%
  filter(!(k9q41 %in%
                  c("No valid response", "Suppressed for confidentiality", "Not in universe", "Logical skip"))) %>%
filter(!(physactiv %in%
                  c("No valid response", "Suppressed for confidentiality", "Not in universe", "Logical skip"))) %>%
filter(!(mold %in%
                  c("No valid response", "Suppressed for confidentiality", "Not in universe", "Logical skip"))) %>%
  rename(.,
        asthma = k2q40a,
        smoke_house = k9q41) %>%
  mutate(.,
         asthma.num = as.numeric(asthma),
         asthma.recode = case_when(
           asthma.num == 2 ~ 0,
           asthma.num == 1 ~ 1
         ))

glimpse(nsch.2.clean)
```

Three operations chained together. The `filter()` calls drop rows where any of our four variables carries a non-response code. The `rename()` gives the two cryptic variables readable names. The `mutate()` builds the outcome.

That last step deserves unpacking, because it is doing something specific. `glmer()` needs a 0/1 numeric outcome, but our factor has levels in the order "Yes", "No", which `as.numeric()` turns into 1 and 2. So `case_when()` maps 1 to 1 and 2 to 0, giving us the 0/1 coding the model requires with 1 meaning the child has asthma.

Always verify a recode rather than assuming it:

```{r}
table(nsch.2.clean$asthma, nsch.2.clean$asthma.recode)
```

A cross-tabulation of the original against the recoded version is the fastest way to confirm the mapping went the direction you intended. Reversed coding here would flip the sign of every coefficient in the chapter.

Note also how many rows the filtering removed. Listwise dropping across four variables adds up, and the analysis sample is what your results describe.

### The `describe` function from `Hmisc` is a nice codebook

```{r}
Hmisc::describe(nsch.2.clean)
```

## Part 2: multilevel logistic regression models

### We use `glmer` instead of `lmer` now that we have a binary outcome

```{r}
model.null <- glmer(asthma.recode ~ (1|fipsst), family = binomial, data = nsch.2.clean)

summary(model.null)
```

Two changes from everything so far: the function is `glmer()` rather than `lmer()`, and we specify `family = binomial`, which tells R the outcome is 0/1 and to use a logit link. Everything else about the syntax is the same.

#### Calculate ICC for logistic regression (L1 random variance is fixed)

```{r}
null.icc <- 0.01808/(0.01808 + (pi^2/3))
null.icc
```

The parenthetical in that heading is the important part. In a linear model we compute the ICC from two *estimated* variances. In a logistic model there is no estimated level-1 variance, because for a binary outcome the variance is determined by the mean. The standard solution fixes it at $\pi^2/3$, roughly 3.29, which is the variance of the standard logistic distribution.

Because that denominator term is assumed rather than estimated, the logistic ICC lives on a latent scale and is not directly comparable to an ICC from a linear model. Treat it as a rough indication of clustering rather than a precise proportion of variance in the observed outcome.

### Let's add a L1 predictor, whether anyone smokes in the house

```{r}
model.2 <- glmer(asthma.recode ~ smoke_house + (1|fipsst), family = binomial, data = nsch.2.clean)

summary(model.2)
```

These estimates are logit coefficients. Long story short, negative numbers suggest lower odds relative to the reference group, and positive numbers suggest higher odds. So here, we would say that children in houses without a smoker have lower odds of having asthma. But that's about all we can say without some more math.

We can exponentiate this coefficient to get an odds ratio, which makes things a little easier to interpret.

```{r}
smoke.odds.ratio <- exp(-0.3877)
smoke.odds.ratio
```

When the OR is less than one (.68 in this case), we subtract 1 − OR which equals .32. This means that for children in houses without a smoker, the odds of having asthma are 32% lower than for children who live in houses with a smoker.

::: {.callout-warning}
## Odds are not probabilities
"32% lower odds" is not the same as "32% less likely," and the two diverge sharply when an outcome is common. Odds are p/(1−p); probability is p. An odds ratio of 0.68 corresponds to different probability differences depending on where you start.

The looser phrasing is only approximately right when the outcome is rare, and it is worth avoiding entirely. This is why the predicted probabilities below matter so much for communicating results.
:::

We can also get predicted probabilities associated with each estimate. To do this, we need to add up the logits and plug them into the formula below:

```{r}
# Predicted probability for child in house with smoker (logit = -1.2380)

exp(-1.2380)/(1+exp(-1.2380))
```

So, for a child in a house with a smoker, the predicted probability of having asthma is 22%.

What about for a child in a house that doesn't have a smoker?

```{r}
# Predicted probability for child in house without smoker (logit = -1.2380 + -0.3877 = -1.6257)

exp(-1.6257)/(1+exp(-1.6257))
```

For a child in a house without a smoker, the predicted probability of having asthma is 16%.

That formula, $e^x/(1+e^x)$, is the inverse logit. It converts a log-odds back into a probability. R has it built in as `plogis()`, so `plogis(-1.6257)` gives the same answer, but writing it out once makes clear what is happening.

Notice the arithmetic in the comment: you build the logit for a profile by *adding up* the relevant coefficients, starting from the intercept. The intercept alone describes a child in the reference category on every predictor.

### Now add another L1 predictor, `mold`

```{r}
model.3 <- glmer(asthma.recode ~ smoke_house + mold + (1|fipsst), family = binomial, data = nsch.2.clean)

summary(model.3)
```

The estimate for `smoke_house` looks very similar (−.35). So the calculations we ran for the first model (ORs, predicted probabilities) would still hold. But we can do the same thing for the `mold` variable.

First, we can exponentiate the coefficient to get an odds ratio:

```{r}
mold.odds.ratio <- exp(-0.4672)
mold.odds.ratio
```

Pretty similar to smoke! So the odds of having asthma for a child in a house without mold (`mold` = No) are about 37 percent lower than the odds for a child in a house with mold.

We can also get predicted probabilities associated with each estimate:

```{r}
# Predicted probability for child in house with mold AND smoker (this is the intercept) (logit = -0.8724)

exp(-0.8724)/(1+exp(-0.8724))
```

So, for a child in a house with a smoker AND mold, the predicted probability of having asthma is 29%.

What about for a child in a house that doesn't have a smoker OR mold?

```{r}
# Predicted probability for child in house without smoker (logit = -0.8724 + -0.3590 + -0.4603 = -1.6917)

exp(-1.6917)/(1+exp(-1.6917))
```

For a child in a house without a smoker OR mold, the predicted probability of having asthma is 15%. Other combinations can be found by adding or removing the logits for each coefficient.

That contrast, 29% against 15%, is the number to lead with when talking to a non-technical audience. It says what fraction of children in each situation we expect to have asthma, in units anyone can act on. No audience outside a statistics seminar will do anything useful with an odds ratio.

### Now add a third L1 predictor, `physactiv`

```{r}
model.4 <- glmer(asthma.recode ~ smoke_house + mold + physactiv + (1|fipsst), family = binomial, data = nsch.2.clean)

summary(model.4)
```

### How about a `mold` by `smoke_house` interaction?

```{r}
model.5 <- glmer(asthma.recode ~ smoke_house + mold + smoke_house:mold + (1|fipsst), family = binomial, data = nsch.2.clean)

summary(model.5)
```

Interactions in logistic models carry an extra wrinkle worth knowing about. Because the logit scale is nonlinear, two predictors can interact on the probability scale even when the interaction coefficient is zero, and vice versa. If the substantive question is about probabilities, check predicted probabilities for each combination rather than relying on the coefficient alone.

### Same thing, PLUS adding a random slope for `mold`

```{r}
model.6 <- glmer(asthma.recode ~ smoke_house + mold + (mold|fipsst), family = binomial, data = nsch.2.clean)

summary(model.6)
```

This asks whether the association between household mold and asthma differs from state to state, which is plausible: climate, housing stock, and building codes all vary.

### Should we keep that random slope? `rand` won't work with GLMMs, but we can compare AIC and BIC

Here, we see that both AIC and BIC went up, which is not a good sign. We probably don't need the random slopes.

```{r}
model.6.aic <- 3045.1
model.6.bic <- 3081.8

model.5.aic <- 3042.7
model.5.bic <- 3073.2

model.6.aic - model.5.aic

model.6.bic - model.5.bic
```

Both differences are positive, meaning the simpler model fits better once complexity is penalized. When AIC and BIC agree, as they do here, the decision is easy. Chapter 6 covers what to do when they disagree.

::: {.callout-tip}
## If `glmer` is slow or will not converge
`glmer()` uses numerical approximation and is considerably more temperamental than `lmer()`. Convergence warnings are common, especially with random slopes. A different optimizer often resolves them:

```r
glmer(y ~ x + (1 | group), family = binomial, data = d,
      control = glmerControl(optimizer = "bobyqa",
                             optCtrl = list(maxfun = 100000)))
```
:::

## Using `modelsummary` and `broom.mixed` to organize your results

```{r}
#| message: false

library(modelsummary)
library(broom.mixed)

models1 <- list(model.null, model.2, model.3, model.4, model.5, model.6)

modelsummary(models1, output = "markdown")
```

By default this table reports log-odds, the raw coefficients. If you want odds ratios instead, add `exponentiate = TRUE`, and label the column clearly so a reader does not assume log-odds.
