# Checking Assumptions for MLMs

**Data for this chapter:** `productivity.dta`

::: {.callout-note}
## Data file needed
Copy `productivity.dta` into this project folder before rendering.
:::

This week, we will use the `productivity` dataset from Stata, developed by Dr. Alicia Munnell and published by the Federal Reserve Bank of Boston (1990). It tracks US states over time, so observations are years nested within states.

Diagnostics are not a pass-fail checklist. They are evidence you weigh, and the useful question is never "was an assumption violated" but "is this violation bad enough to change what I conclude." This chapter runs the standard checks and then does the thing that actually settles the question: refits the model under different assumptions and sees whether the answer moves.

## Load in our MVP packages

```{r}
#| label: setup
#| message: false

suppressPackageStartupMessages(library(tidyverse))
library(lme4)
library(Hmisc)
```

## Part 1: loading in the data, and then recoding variables

### Load in the data

```{r}
productivity <- haven::read_dta("productivity.dta")

glimpse(productivity)
```

### The `describe` function from `Hmisc` is a nice codebook

```{r}
Hmisc::describe(productivity)
```

## A little quick cleaning

Convert region to a factor variable, rescale public capital stock, "zero out" year, and create nonlinear terms.

```{r}
prod.clean <- productivity %>%
  mutate(.,
         region.fac = as_factor(region),
         year0 = year - 1970,
         year_sq = year0^2,
         year_cube = year0^3,
         pub_cap_1000 = public/1000)

glimpse(prod.clean)
```

Two of these deserve a note.

Subtracting 1970 from `year` puts the intercept at the start of the observation window rather than at year zero of the common era, which is both meaningless and thousands of units away from any data we have. Polynomial terms built from the recentered variable are also far less collinear with each other than they would be otherwise.

Dividing public capital by 1,000 is rescaling, and it changes nothing about the model's fit. What it changes is readability: a coefficient expressed per thousand dollars is easier to report than one carrying several leading zeros. Rescaling a predictor is always safe. Rescaling only *some* predictors and forgetting which is where people get into trouble, so a name like `pub_cap_1000` that records the units is worth the extra characters.

## Part 2: find your "final" or preferred model

### Start with a null growth model since this is longitudinal data

```{r}
model.null.growth <- lmer(unemp ~ year0 + (1|state), REML = FALSE, data = prod.clean)

summary(model.null.growth)
```

The outcome is state unemployment rate, and the clustering variable is state. This is the same structure as the growth models from Module 8, with states in place of children and years in place of grades.

### Add in quadratic effects

```{r}
model.2 <- lmer(unemp ~ year0 + year_sq + (1|state), REML = FALSE, data = prod.clean)

summary(model.2)
```

### Add in cubic effects

```{r}
model.3 <- lmer(unemp ~ year0 + year_cube + (1|state), REML = FALSE, data = prod.clean)

summary(model.3)
```

### Both quadratic and cubic effects

```{r}
model.4 <- lmer(unemp ~ year0 + year_sq + year_cube + (1|state), REML = FALSE, data = prod.clean)

summary(model.4)
```

Unemployment over the 1970s and 1980s went through recessions and recoveries, so a straight line is unlikely to describe it well. The polynomial terms give the trajectory room to bend more than once.

### Now for some more predictors

```{r}
model.5 <- lmer(unemp ~ year0 + year_sq + year_cube + gsp + private + pub_cap_1000 + (1|state), REML = FALSE, data = prod.clean)

summary(model.5)
```

This is our "final" model, and everything that follows examines it rather than building further.

## Use the `augment` function from the `broom.mixed` package

This gets predictions, residuals, Cook's distances, and more.

```{r}
#| message: false

library(broom.mixed)

diagnostics <- augment(model.5)
```

`augment()` returns the original data plus a set of new columns whose names begin with a dot: `.fitted`, `.resid`, `.cooksd`, and others. Everything in this chapter works from that data frame.

One thing worth knowing: `diagnostics` contains one row per observation *the model actually used*. If any rows were dropped for missing data, this frame is smaller than `prod.clean`. That is a feature, since it means residuals and fitted values always line up, but it does mean you should use `diagnostics` rather than the original data for anything downstream.

## Step 1: assess normality of residuals

### Visually, with a histogram

```{r}
#| label: fig-resid-hist
#| fig-cap: "Histogram of residuals with reference lines at ±2.5."

ggplot(data = diagnostics, mapping = aes(x = .resid)) +
  geom_histogram(binwidth = .25) + theme_classic() +
  labs(title = "Histogram of Residuals for State Unemployment Model",
                      x = "Residual Value") +
  geom_vline(xintercept = c(-2.5, 2.5), linetype = "dotted")
```

The dotted reference lines mark ±2.5, a rough boundary beyond which residuals are worth a closer look. What you are checking for is severe skew, heavy tails, or obvious multimodality, not perfect symmetry.

### Statistically, with the Shapiro-Wilk test

```{r}
shapiro.test(diagnostics$.resid)
```

Expect this to be significant, and do not make much of it. Shapiro-Wilk detects arbitrarily small departures from normality once the sample is reasonably large, and with several hundred observations it will reject for essentially any real dataset.

The histogram carries more information than the test. It is also worth remembering what this assumption buys: normality of residuals matters mainly for small-sample inference, and at these sample sizes the central limit theorem does most of the work regardless.

::: {.callout-note}
## A size limit on this test
`shapiro.test()` accepts between 3 and 5,000 observations and errors outside that range. If you apply it to a larger model, sample first:

```r
set.seed(664)
shapiro.test(sample(diagnostics$.resid, 5000))
```
:::

## Step 2: use a residuals vs. fitted (RVF) plot to assess homoskedasticity of errors

### Basic RVF plot, without groups

```{r}
#| label: fig-rvf
#| fig-cap: "Residuals versus fitted values."

ggplot(data = diagnostics, mapping = aes(x = .fitted, y = .resid)) +
  geom_point() + labs(title = "RVF Plot for State Unemployment Model",
                      x = "Predicted Value, % Unemployment",
                      y = "Residual Value") + theme_classic()
```

Read this plot for shape rather than for scatter. A fan or funnel means the residual variance depends on the fitted value, which is heteroskedasticity. A curve means the relationship is not linear in the predictors as specified. An even, shapeless cloud is what you want.

Be specific in your write-up about *where* in the fitted range any problem appears. A violation confined to a sparse tail is a different matter from one across the whole range.

### Same thing, but with added color by year0

```{r}
#| label: fig-rvf-year
#| fig-cap: "The same plot, colored by year."

ggplot(data = diagnostics, mapping = aes(x = .fitted, y = .resid, color = year0)) +
  geom_point() + labs(title = "RVF Plot for State Unemployment Model, by Year",
                      x = "Predicted Value, % Unemployment",
                      y = "Residual Value") + theme_classic()
```

Adding color turns a general diagnostic into a specific one. If residuals cluster by year, with all the dark points high and the light ones low, the model is systematically missing something about time. That is far more actionable than knowing the residuals are merely uneven, and it is the kind of pattern longitudinal data produces routinely.

## Step 3: use Cook's distance to identify outliers

```{r}
#| label: fig-cooks
#| fig-cap: "Cook's distance by fitted value, labelled by state."

ggplot(data = diagnostics, mapping = aes(x = .fitted, y = .cooksd, label = state)) +
  geom_point() + geom_text(nudge_x = .25) + theme_classic() +
  labs(title = "Cook's Distance Plot for State Unemployment Model",
                      x = "Predicted Value, % Unemployment",
                      y = "Cook's Distance") +
  geom_hline(yintercept = 4/816, linetype = "dotted")
```

Labelling the points is what makes this plot useful rather than decorative. An unlabelled outlier is a puzzle; a labelled one is a case you can go look at.

The reference line at 4/816 is the conventional 4/n cutoff, where 816 is the number of observations. It is a flag, not a verdict.

When a case is flagged, look at it before doing anything. An observation with an impossible value is a data error and should be fixed or removed. An observation that is simply unusual, a state with a genuinely severe recession, is data. Deleting inconvenient observations because a statistic flagged them is not a diagnostic procedure; it is a way of guaranteeing the result you already wanted.

## Step 4: re-run model with robust standard errors

```{r}
#| message: false

library(robustlmm)

model.robust <- rlmer(unemp ~ year0 + year_sq + year_cube + gsp + private + pub_cap_1000 + (1|state), REML = FALSE, data = prod.clean)

summary(model.robust)
```

`rlmer()` from `robustlmm` fits the same model but downweights observations with large residuals rather than letting them exert full influence. If the estimates barely move relative to `model.5`, outliers were not driving your results.

This model can take a while to fit, since it works iteratively.

## Step 5: re-run model with outliers removed (Cook's distance > .10)

```{r}
prod.trimmed <- diagnostics %>%
  filter(., .cooksd < .10)

model.trimmed <- lmer(unemp ~ year0 + year_sq + year_cube + gsp + private + pub_cap_1000 + (1|state), REML = FALSE, data = prod.trimmed)

summary(model.trimmed)
```

Note that the filtering happens on `diagnostics`, not on `prod.clean`. That works because `augment()` kept all the original columns alongside the new diagnostic ones, so the trimmed frame still has everything the model formula needs.

```{r}
nrow(diagnostics); nrow(prod.trimmed)
```

Reporting how many observations came out is not optional. A reader needs to know whether the robustness check dropped three cases or three hundred.

## Step 6: compare original "final" model, robust model, and trimmed model

This is the sensitivity analysis.

```{r}
#| message: false

library(modelsummary)

models <- list(model.5, model.trimmed)

modelsummary(models, output = "markdown")
```

This is the part that actually settles the question. Rather than arguing about whether assumptions hold, refit under different assumptions and compare.

Read across the columns for the coefficient you care about most. If it holds steady in magnitude, direction, and significance, you have an answer to the reviewer worried about your residuals: the conclusion does not depend on the assumption in question. That is a much stronger response than asserting the assumption was met.

## Writing it up

A sensitivity analysis paragraph belongs in nearly every quantitative paper, and it has a predictable shape. State which diagnostics you examined. Report what they showed, honestly, including the parts that looked imperfect. Describe the robustness checks you ran. Then state whether your substantive conclusion held.

The version that reads as defensive says the assumptions were met and moves on. The version that reads as competent concedes that residuals showed mild skew and a handful of state-years exceeded the conventional Cook's distance threshold, then reports that the coefficients changed negligibly under robust estimation and with those cases excluded. The second is more honest and considerably more persuasive.
